Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-05-24 doc/mc-paper.tex
Minor tweaks

Capelli, Lagniez, and Marquis developed a knowledge compiler that
generates a certificate in a proof system that is itself based on
dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their \cdfour{}
program, a modified version of \dfour{}, generates annotations to the
compiled representation, providing information about how the compiled
version relates to the input clauses.  It also generates a file of
clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
Completing the certification involves running two different checkers
on the annotated dec-DNNF graph and the DRAT file.  Although the
authors make informal arguments regarding the soundness of their
frameworks, these do not provide strong levels of assurance.  Indeed,
we have identified a weakness in the their methodology due to an
invalid assumption about the guarantees provided by \dtrim{}, the
program it uses to check the DRAT file.  Moreover, this weakness is
\emph{exploitable}: their framework can be ``spoofed'' into accepting
an incorrect compilation.

In more detail, \cdfour{} emits a sequence of clauses $R$ that
includes the conflict clauses that arose via a top-down processing of
the input clauses.  Given input formula $\phi_I$, their first task is to
check whether $\phi_I \imply R$, i.e., that any assignment that satisfies $\phi_I$ also satisfies each of the clauses in $R$.
They then base other parts of their proof on that
property and use a separate program to perform a series of additional
checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
satisifies the \emph{reverse asymmetric tautology} (RAT) property with
respect to the preceding
clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
clause $C$ to a set of clauses maintains satisfiability, but it may
not preserve models.  As an example, consider the following formulas:
\begin{center}
  \begin{tabular}{lccc}
    $\phi_1$: & $(x_1 \lor x_3)$ & & \\
    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_2 \lor \obar{x}_3)$\\
  \end{tabular}
\end{center}
Clearly, these two formulas are not equivalent---$\phi_1$ has six
models, while $\phi_2$ has four.  In particular, $\phi_1$ allows
arbitrary assignments to variable $x_2$.  Critically, however, the
second clause of $\phi_2$ is RAT with respect to $\phi_1$---any
satisfying assignment to $\phi_1$ can be transformed into one that
also satisfies $\phi_2$ by setting $x_2$ to 1, while keeping the values for other variables fixed.

@@

This weakness can be corrected by restricting \dtrim{} to only add
clauses that obey the stronger \emph{reverse unit propagation} (RUP) property~\cite{goldberg,vangelder08_verifying_rup_proofs}.  We have added a
command-line argument to \dtrim{} that enforces this
restriction.\footnote{Available at
\url{https://github.com/marijnheule/drat-trim/releases/tag/v05.22.2023}.}  This weakness, however, illustrates the general challenge of
developing a new proof framework.
As we can attest,
without engaging in the effort to formally verify the framework, there are likely to be
conditions that make the framework unsound.

@@

The declaration of a \emph{product} operation, creating a node with operation $\pand$,
 has the form:
\begin{center}
\begin{tabular}{ccccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{\texttt{p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} &
  \makebox[5mm]{$\cdots$} & \makebox[5mm]{$\lit_k$} & \makebox[5mm]{\texttt{0}} \\
\end{tabular}
\end{center}
Integer $i$ is a new clause ID, $v$ is a positive integer that does not
correspond to any previous variable, and $\lit_1, \lit_2, \ldots, \lit_k$ is a sequence of $k$
integers, indicating the arguments as literals of existing variables.
As Table~\ref{tab:defining}(A) shows,
this declaration implicitly causes $k+1$ clauses to be added to the proof, using a Tseitin encoding to define extension variable $v$ as the product of its arguments.

@@

The declaration of a \emph{sum} operation, creating a node with operation $\por$, has the form:
\begin{center}
\begin{tabular}{ccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{\texttt{s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$}
\makebox[5mm]{$H$} & \makebox[5mm]{$\texttt{0}$} \\
\end{tabular}
\end{center}
Integer $i$ is a new clause ID, $v$ is a positive integer that does
not correspond to any previous variable, and $\lit_1$ and $\lit_2$ are
signed integers, indicating the arguments as literals of existing variables.  Hint $H$
consists of a
sequence of clause IDs, all of which must be defining clauses for other POG operations.\footnote{The restriction to defining clauses in the hint is critical to soundness.
Allowing the hint to include the IDs of input clauses creates an exploitable weakness.  We discovered this weakness in the course of our efforts at formal verification.}
As Table~\ref{tab:defining}(B) shows,
this declaration implicitly causes three clauses to be added to the proof, using a Tseitin encoding to define extension variable $v$ as be the sum of its arguments.
The hint must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$, showing that the two children of this node have disjoint models.

@@

In comparing other proof frameworks, we found that the \cdfour{}
toolchain ran very fast and could handle very large benchmarks.  Even
with a total time limit of 1,000 seconds, including the time for
knowledge compilation, the \cdfour{} toolchain completed 106
benchmarks, while the CPOG toolchain completed just 82.  We found
that the runtimes for the MICE toolchain versus our CPOG toolchain
showed little correlation, reflecting the fact that the two solve
different problems and use different approaches.  In general, our CPOG
toolchain showed better scaling, in part due to its ability to control
the recursion through lemmas.  On the other hand, neither of these
toolchains could perform end-to-end verification when the knowledge
compilation was preceded by preprocessing.

>>>>

Capelli, Lagniez, and Marquis developed a knowledge compiler that
generates a certificate in a proof system that is itself based on
dec-DNNF~\cite{capelli:sat:2019,capelli:aaai:2021}.  Their \cdfour{}
program, a modified version of \dfour{}, generates annotations to the
compiled representation, providing information about how the compiled
version relates to the input clauses.  It also generates a file of
clausal proof steps in the DRAT format~\cite{wetzler14_drattrim}.
Completing the certification involves running two different checkers
on the annotated dec-DNNF graph and the DRAT file.  Although the
authors make informal arguments regarding the soundness of their
frameworks, these do not provide strong levels of assurance.  Indeed,
we have identified a weakness in their methodology due to an
invalid assumption about the guarantees provided by \dtrim{}, the
program it uses to check the DRAT file.  This weakness is
\emph{exploitable}: their framework can be ``spoofed'' into accepting
an incorrect compilation.

In more detail, \cdfour{} emits a sequence of clauses $R$ that
includes the conflict clauses that arose via a top-down processing of
the input clauses.  Given input formula $\phi_I$, their first task is to
check whether $\phi_I \imply R$, i.e., that any assignment that satisfies $\phi_I$ also satisfies each of the clauses in $R$.
They then base other parts of their proof on that
property and use a separate program to perform a series of additional
checks.  They use \dtrim{} to prove the implication, checking that each clause in $R$
satisifies the \emph{reverse asymmetric tautology} (RAT) property with
respect to the preceding
clauses~\cite{heule:cade:2013,jarvisalo:ijcar:2012}.  Adding a RAT
clause $C$ to a set of clauses maintains satisfiability, but does not necessarily
preserve models.  As an example, consider the following formulas:
\begin{center}
  \begin{tabular}{lccc}
    $\phi_1$: & $(x_1 \lor x_3)$ & & \\
    $\phi_2$: & $(x_1 \lor x_3)$ & $\land$ & $(x_2 \lor \obar{x}_3)$\\
  \end{tabular}
\end{center}
Clearly, these two formulas are not equivalent---$\phi_1$ has six
models, while $\phi_2$ has four.  In particular, $\phi_1$ allows
arbitrary assignments to variable $x_2$.  Critically, however, the
second clause of $\phi_2$ is RAT with respect to $\phi_1$---any
satisfying assignment to $\phi_1$ can be transformed into one that
also satisfies $\phi_2$ by setting $x_2$ to 1, while keeping the values for other variables fixed.

@@

This weakness can be corrected by restricting \dtrim{} to only add
clauses that obey the stronger \emph{reverse unit propagation} (RUP) property~\cite{goldberg,vangelder08_verifying_rup_proofs}.  We have added a
command-line argument to \dtrim{} that enforces this
restriction.\footnote{Available at
\url{https://github.com/marijnheule/drat-trim/releases/tag/v05.22.2023}.}  This weakness, however, illustrates the general challenge of
developing a new proof framework.
As we can attest,
without engaging in an effort to formally verify the framework, there are likely to be
conditions that make the framework unsound.

@@

The declaration of a \emph{product} operation, creating a node with operation $\pand$,
 has the form:
\begin{center}
\begin{tabular}{ccccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{\texttt{p}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$} &
  \makebox[5mm]{$\cdots$} & \makebox[5mm]{$\lit_k$} & \makebox[5mm]{\texttt{0}} \\
\end{tabular}
\end{center}
Integer $i$ is a new clause ID, $v$ is a positive integer that does not
correspond to any previous variable, and $\lit_1, \lit_2, \ldots, \lit_k$ is a sequence of $k$
integers, indicating the arguments as literals of existing variables.
As Table~\ref{tab:defining}(A) shows,
this declaration implicitly causes $k+1$ clauses to be added to the proof, providing a Tseitin encoding that defines extension variable $v$ as the product of its arguments.

@@

The declaration of a \emph{sum} operation, creating a node with operation $\por$, has the form:
\begin{center}
\begin{tabular}{ccccccc}
  \makebox[5mm]{$i$} & \makebox[5mm]{\texttt{s}} & \makebox[5mm]{$v$} & \makebox[5mm]{$\lit_1$} & \makebox[5mm]{$\lit_2$}
\makebox[5mm]{$H$} & \makebox[5mm]{$\texttt{0}$} \\
\end{tabular}
\end{center}
Integer $i$ is a new clause ID, $v$ is a positive integer that does
not correspond to any previous variable, and $\lit_1$ and $\lit_2$ are
signed integers, indicating the arguments as literals of existing variables.  Hint $H$
consists of a
sequence of clause IDs, all of which must be defining clauses for other POG operations.\footnote{The restriction to defining clauses in the hint is critical to soundness.
Allowing the hint to include the IDs of input clauses creates an exploitable weakness.  We discovered this weakness in the course of our efforts at formal verification.}
As Table~\ref{tab:defining}(B) shows,
this declaration implicitly causes three clauses to be added to the proof, providing a Tseitin encoding that defines extension variable $v$ as the sum of its arguments.
The hint must provide a RUP proof of the clause $\obar{\lit}_1 \lor \obar{\lit}_2$, showing that the two children of this node have disjoint models.

@@

In comparing other proof frameworks, we found that the \cdfour{}
toolchain ran very fast and could handle very large benchmarks.  Even
with a total time limit of 1,000 seconds, including the time for
knowledge compilation, the \cdfour{} toolchain completed 106
benchmarks, while the CPOG toolchain completed just 82.  We found
that the runtimes for the MICE toolchain versus our CPOG toolchain
showed little correlation, reflecting the fact that the two solve
different problems and use different approaches.  In general, our CPOG
toolchain showed better scaling, in part due to its ability to control
the recursion through lemmas.  Neither of these prior
toolchains could perform end-to-end verification when the knowledge
compilation was preceded by preprocessing.
